---
title: "Reading ISRIC Phosphorus Retention Data"
output:
  html_notebook:
    highlight: tango
    number_sections: true
    theme: spacelab
    toc: true
    toc_float: true
---

# Loading Libraries

```{r, message=FALSE, warning=FALSE, results="hide"}
library(raster)
library(soilP)
library(dplyr)
library(tidyr)
library(Hmisc)
library(MASS)
library(ggplot2)
library(ade4)
```
# Reading ISRIC 2011 data
## Starting with the base raster file
First we load the raster. 
```{r}
data_dir    <- system.file("data",    package = "soilP", mustWork = TRUE)
extdata_dir <- system.file("extdata", package = "soilP", mustWork = TRUE)

FAO74_tif <- file.path(extdata_dir,"ISRIC2011", "FAO74.tif")

# Initializing ISRIC2011 list with the base map
ISRIC2011 <- list()
ISRIC2011$FAO74 <- raster::raster(FAO74_tif)

```

This raster has a single layer and its  value corresponds to the soil map unit 
identifier (ID) for the Soil Map of the World, FAO74. The same identifier can
be found in more recent versions of the FAO soil map and in the harmonized soil 
database.

## Plotting FAO Soil Map Unit Raster
```{r, fig.height=4, fig.width=10, warning=FALSE}
nb_plot(ISRIC2011$FAO74, main = "Soil Map of the World, FAO74")
```

The MUIDs have the lowest values in Africa and then ascend through Asia,
the Americas, Oceania and attain the highest values in Europe.
Contrary to what the file name seems to imply these ids do not directly
represent  the phosphorus retention potential. Batjes used the ids and 
a series of SQL queries to assign phosphorus retention potential classes
to each map unit, which then are easily projected over each pixel in the
map.

These main classes can be extracted from the map unit raster, and the tables in
the mdb files. These tables function as raster attribute tables.

## Extracting Raster Attribute Table from Microsoft Access database (`mdb` file).

For this I need Hmisc and the ODBC drivers. Mac installation is from brew.

The MS Access database has the tables for mapping the soil unit identifiers to
soil properties, i.e. phosphorus retention as soil unit composition percentage, 
or physicochemical variables.

```{r}

mdb_file <- file.path(
  extdata_dir,
  "ISRIC2011",
  "ISRIC_Phosphorus_Retention_Potential.mdb")
```

## Adding Raster Attribute Table, including Map Unit P Retention Potential 

`levels<-` S3 method is internal to `raster` and it could not be
exported  to `soilP`, which prevented me from making a working package function
out of the following snippet:


```{r}

### Get Raster Attribute Tables
ISRIC_AT <- list()

ISRIC_AT <- read_ISRIC_AT(mdb_file)

### Adding ID Raster Attribute Table to FAO74
ISRIC2011$FAO74 <- raster::raster(FAO74_tif)
ISRIC2011$FAO74 <- raster::ratify(ISRIC2011$FAO74)

levels(ISRIC2011$FAO74) <- ISRIC_AT$mapunit

```
# Soil Map Phosphorus Retention Potential as Percentages of Map Units

A phosphorus retention potential class was assigned to each of the FAO74
soil units (Low, Moderate, High and very High). 
Each map unit is composed of up to 8 soil units in different area percentages.
Depending on the composition of the map unit and each component soil unit 
phosphorus retention potential class a  map unit class was assigned.
In total 260 different phosphorus retention potential classes were assigned to
the 4923 map units. These were further grouped into 16 main phosphorus retention
potential classes.

```{r, fig.height=4, fig.width=10, warning=FALSE}

pct_vars <- c("Lo", "Mo", "Hi", "VH", "MISC") 

ISRIC2011 <- c(ISRIC2011, layers_from(ISRIC2011$FAO74,cols = pct_vars))

for (varname in pct_vars) {
  out_tif <- file.path( extdata_dir, "ISRIC2011", paste0(varname,".tif"))
    
  writeRaster(ISRIC2011[[varname]],
              file = out_tif,
              datatype = 'INT1U',
              overwrite = TRUE)
    
  nb_plot(ISRIC2011[[varname]], main =  varname)
}


```

# Rebuilding the arcgis render appeareance 

We will use a manually curated raster attribute table stored in th `soilclass`
dataframe.

See the manual for details.
```{r}
?soilclass
```

## Adding Raster Attribute Table to  `ISRIC2011$FAO74`

First we need to  merge `soilclass` to the Soil Map Unit id number, then we add this as the RAT to the FAO74 raster.


```{r}

### Adding ID Raster Attribute Table to FAO74

ISRIC_AT$mu_soilclass <- ISRIC_AT$mapunit %>% 
    dplyr::left_join(soilclass, by = c("mainclass" = "main"))

levels(ISRIC2011$FAO74) <- ISRIC_AT$mu_soilclass
```

##  Reclassifying the FAO74 soil map into ascending Phosphorus Retention Classes

I manually assigned integers in ascending order to the
`ascending` column in the `soilclass` dataframe as follows: For the Low map unit
main class, higher percentage of Low soil unit P retention corresponded to lower
integers. For the Moderate class, a multivariate analysis reveled that Mo2 has a
higher content of Low souil units,snd  simirlaly Mo3 has a greater content of Hi
soil units tan Mo1, which suggests a Mo2, Mo1, Mo3, ascending order.
For the High, and very High map unit main classes, higher integers were assigned
to higher percentage of corresponding soil unit P retention. 
This means the higher the Low soil unit P retention percentage the lower 
the integer, and complentarily the higher the High, and Very 
High soil unit P retention percentage the higher the integer, except in the
Moderate class that behaves unexpectedly.


```{r}

ISRIC2011$main <- layers_from(ISRIC2011$FAO74, cols = "ascending")[[1]]

raster::writeRaster(
  ISRIC2011$main,
  file.path(extdata_dir,"ISRIC2011","main.tif"),
  datatype = 'INT1U',
  overwrite = TRUE)

```

## Plotting Soil Phosphorus Retention Potential Raster

```{r, fig.height=4, fig.width=10, warning=FALSE}

nb_plot(ISRIC2011$main,
        axis.args = list(breaks=0:15, at = 0:15, cex.axis = 2),
        main =  "Phosphorus Retention Potential, ISRIC2011")

```

Now the numbers do represent the phosphorus retention potential!!!

However, the `raster` plot legend above assumes a continuous scale from 0 to 15,
while the data is explicitly a categorical variable (although derived from
continuous percentages, see above). 


```{r}
# Raster Attribute Table
ISRIC2011$main <- raster::ratify(ISRIC2011$main)
# Assign RAT to main class raster
levels(ISRIC2011$main) <- rat(ISRIC2011$main) %>% 
    dplyr::inner_join(soilclass, by = c("ID" = "ascending"))
```

## Adding arcgis render color and saving the map as full resolution geotif

```{r, fig.height=4, fig.width=10, warning=FALSE}

main_color <- layers_from(ISRIC2011$FAO74, cols = c("r","g","b"))

plotRGB(stack(main_color))

writeRaster(stack(main_color),
  filename  = file.path(extdata_dir, "ISRIC2011","main_color.tif"),
  datatype  = "INT1U",
  options   = "TFW=YES",
  format    = "GTiff",
  overwrite = TRUE)
```

## Using rasterVis to appropiately label categories and add legend 
Remember that we assigned to the to the phosphorus retention potential main
classes an *ad hoc* order so we can interpret a direction of ascending retention 
potential (see above) in the plot legend.

In order to show the right labels in this ascending scale we use the `levelplot`
function from `RasterVis` that uses the raster attribute table to transform the 
numerical value of the raster to discrete categories.


```{r}
rasterVis::levelplot(
  ISRIC2011$main, att = "main",
  col.regions = soilclass$color_hex[1:16],
  maxpixels = ncell(ISRIC2011$main),
  scales = list(draw = FALSE),
  xlab = NULL, ylab = NULL,
  main = "Generalized Phosphorus Retention Potential Map")

```



# Multivariate analysis of Soil retention Potential

We should establish a data based order of map unit soil phosphorus 
retention classes instead of postulating an ad hoc order. This can be done
through selecting the first discriminant dimension from a Discriminant
Analysis of the percentage of soil unit phosphorus retention classes per
mapping unit. Furthermore, we can use that discriminant dimention as a
continuous variable instead of the discrete classes for downstream
analyses
      
## Map Unit Soil Composition Matrix and entropy

Each soil unit percentage can be used as a map unit descriptor variable for 
multivariate analysis of the phosphorus retention potential per map unit.

```{r}
ISRIC_AT <- within(ISRIC_AT, {
  su_share <- soil_composition(mapunit,FAO74)
  su_share <- as.data.frame(
    cbind( ID = mapunit$ID,
           su_share,
           soilS = row_entropy(su_share/100)))
})

```
However this results in sparse matrix with essentially orthogonal columns.

## Phosphorus Retention, Soil Unit Composition  Join table

```{r}
phychem_vars <- c("pH","SAND","SILT","CLAY","CECLAY")
su_types <- colnames(ISRIC_AT$su_share)
ret_vars <- c("ID",
              "mainclass",
              phychem_vars,
              pct_vars,
              su_types)

levels <- soilclass %>%
            dplyr::arrange(ascending) %>%
            dplyr::select(main)

ISRIC_AT <- within(ISRIC_AT, {
  ret_share <- mapunit %>%
    dplyr::left_join(soilunit, by = c("SOIL1" = "key")) %>%
    dplyr::left_join(su_share, by = "ID") %>%
    dplyr::select(!!!ret_vars) 
  ret_share$mainclass <- factor(ret_share$mainclass, levels = levels$main[1:16])
})


```

## Soil and Phosphorus Retention Class Composition Entropy
```{r}
ISRIC_AT <- within(ISRIC_AT, {
  ret_share$pS <- row_entropy(ret_share[,pct_vars] / 100)
})

with(ISRIC_AT,{
  boxplot(soilS ~ mainclass, data = ret_share, las = 2, 
          main = "Map Unit Soil Composition Entropy")
  boxplot(pS ~ mainclass, data = ret_share, las = 2,
          main = "Phosphorus Retention Class Composition Entropy")
})
```

## Exploratory Discriminant Analysis using `ade4`

Soil Retention Potential Space is essentially a tetrahedron with Moderate
retention at its center.Only three degrees of freedom, Lo-VH, Mo-Hi, Lo-Hi.


```{r}
# create dudi object for discriminant analysis

soil_idx <- with(ISRIC_AT$ret_share,{
    which(is_soil(mainclass))
})

soil_data <- ISRIC_AT$ret_share[soil_idx,-1]

soil_data$mainclass   <- with(ISRIC_AT$ret_share,{
    factor(mainclass[soil_idx], levels = levels$main[5:16])
})



pca <- dudi.pca(soil_data[,-1], scannf = FALSE, nf = 139)
  
dis <- discrimin(pca, fac = soil_data$mainclass, scannf = FALSE, nf = 10)
palette <- soilclass$color_hex[5:16]
color <- palette[soil_data$mainclass]

# Select most important variables
disva <- as.data.frame(dis$va)
disva$idx <- (1:nrow(dis$va)) / 1000

major_vars <- 1000 * disva %>% 
  dplyr::filter_all(any_vars(abs(.) > 0.5)) %>%
  dplyr::select(idx)

# Plot
par(mfrow = c(2, 2))
s.class(dis$li, xax = 2, yax = 1, fac = soil_data$mainclass, col = palette)
s.class(dis$li, xax = 3, yax = 1, fac = soil_data$mainclass, col = palette)
s.class(dis$li, xax = 3, yax = 2, fac = soil_data$mainclass, col = palette)
s.corcircle(dis$va[major_vars$idx,])
par(mfrow = c(1, 1))
```


# Linearization of Phosphorus Retention Potential with `MASS` LDA

The `MASS` LDA wiill use all info by default.

```{r}

fit <- MASS::lda(mainclass ~ ., data = soil_data)

plda <- predict(object = fit, # predictions
                 newdata = ISRIC_AT$ret_share[,-c(1,2)])

```

From categorical variable to continous "linear" scale.
Linear combination of map unit soil content.

```{r}
ret_scale<- NA
ISRIC_AT <- within(ISRIC_AT,{
  ret_scale <- ret_share[,c("ID",pct_vars)]
  ret_scale$ret <- ret_share$mainclass
  ret_scale[!is_soil(ret_scale$ret), pct_vars] <- NA
})


ISRIC_AT$ret_scale <- within(ISRIC_AT$ret_scale,{
    # Weighted scale, ad hoc weights!!!!!!    
    weighted <- as.double((Lo + 2 * Mo + 3 * Hi + 4 * VH) / 400)
    # Linear Discrimant scale, post hoc weights/coefficients
    LD1 <-  plda$x[,1]
    LD2 <-  -plda$x[,2] # sign manually adjusted
    LD3 <-  -plda$x[,3] # sign manually adjusted
    LD1[!is_soil(ret)] <- NA
    LD2[!is_soil(ret)] <- NA
    LD3[!is_soil(ret)] <- NA
})

```

```{r, fig.asp = 1}
with(ISRIC_AT,{
  plot_P_scales(ret_scale[soil_idx,], palette = palette,
              scale_x = "LD2",scale_y = "weighted")
})

```

```{r}
ld_vars <- c("weighted", "LD1", "LD2", "LD3")

ISRIC_AT <- within(ISRIC_AT,{
  raster_scale <- mu_soilclass[,c("ID",pct_vars)] %>% 
    dplyr::left_join(ret_scale[,c("ID",ld_vars)],
                     by = c("ID" = "ID")) %>%
    dplyr::mutate_if(colnames(.) %in% ld_vars, scale256)
})
```

## Linear Discriminant Maps
```{r, fig.height=4, fig.width=10, warning=FALSE}

levels(ISRIC2011$FAO74) <- ISRIC_AT$raster_scale

ISRIC2011 <- c(ISRIC2011, layers_from(ISRIC2011$FAO74, cols = ld_vars))

for(varname in ld_vars) {

  out_tif <- file.path( extdata_dir, "ISRIC2011", paste0(varname,".tif"))
    
  writeRaster(ISRIC2011[[varname]],
              file = out_tif,
              datatype = 'INT1U',
              overwrite = TRUE)
    
  nb_plot(ISRIC2011[[varname]], main =  varname)
}

# finally save ISRIC2011 in .RData format
save(ISRIC_AT, file = file.path(data_dir, "ISRIC_AT.RData"))
save(ISRIC2011, file = file.path(data_dir, "ISRIC2011.RData"))
```

## False Color Combination of Linear Discriminants
```{r, fig.height=4, fig.width=10, warning=FALSE}

LD_composite <- stack(ISRIC2011[c("LD2","LD1","LD3")])

op <- par()
par(mar = c(0, 0, 0, 0), oma = c(0, 0, 0, 0))

plotRGB(LD_composite)

par(op)

out_tif <- file.path(extdata_dir,"ISRIC2011","LD_composite.tif")
writeRaster(LD_composite, filename = out_tif,
            datatype  = "INT1U",
            options   = "TFW=YES",
            format    = "GTiff",
            overwrite = TRUE)
```
